### 2020-07-26 Model Data Sheets and Code/Zaca MCF Mortality Model Results.Rmd
minor changes

mcf <- read.csv('Zaca MCF variables for model.csv') %>% 
  mutate(mortality_qual= fct_relevel(mortality_qual, levels=c("High", "Moderate", "Low", "None"))) %>% 
  mutate(dnbr_qualitative = case_when(
    dnbr_2007 < 0.1 ~ "unburned",
    dnbr_2007 < 0.27 ~ "low",
    dnbr_2007 < 0.66 ~ "moderate",
    dnbr_2007 > 0.66 ~ "high")) %>%
  mutate(dnbr_qualitative = as.factor(dnbr_qualitative),
         dnbr_qualitative = fct_relevel(dnbr_qualitative, levels=c("high", "moderate", "low", "unburned"))) %>%
  mutate(aspect_transformed = (1-cos((2*pi*aspect)/360))/2) %>% 
  mutate(three_class_mortality = case_when(
    mortality_qual %in% "High" ~ "High",
    mortality_qual %in% "Moderate" ~ "Moderate",
    mortality_qual %in% c("Low", "None") ~ "Low")) %>% 
  mutate(three_class_mortality = as.factor(three_class_mortality)) %>% 
  mutate(three_class_mortality= fct_relevel(three_class_mortality, 
                                     levels=c("Low", "Moderate", "High"))) %>% 
  mutate(two_class_mortality = case_when(
    mortality_qual %in% "High" ~ "High",
    mortality_qual %in% c("Moderate", "Low", "None") ~ "Other")) %>% 
  mutate(two_class_mortality = as.factor(two_class_mortality)) %>% 
  mutate(two_class_mortality= fct_relevel(two_class_mortality, 
                                     levels=c("Other", "High"))) %>% 
  mutate(tc_mortality = as.factor(tc_mortality)) %>% 
  mutate(rdnbr_qualitative = case_when(
    rdnbr_2007 < 69 ~ "None",
    rdnbr_2007 < 315 ~ "Low",
    rdnbr_2007 < 640 ~ "Moderate",
    rdnbr_2007 > 640 ~ "High")) %>%
  mutate(rdnbr_qualitative = as.factor(rdnbr_qualitative),
         rdnbr_qualitative = fct_relevel(rdnbr_qualitative, 
                                         levels=c("High", "Moderate", "Low", "None"))) %>% 
    mutate(rdnbr_qualitative2 = case_when(
    rdnbr_qualitative %in% "High" ~ "High",
    rdnbr_qualitative %in% c("Moderate", "Low", "None") ~ "Other")) %>% 
  mutate(rdnbr_qualitative2 = as.factor(rdnbr_qualitative2)) %>% 
  mutate(rdnbr_qualitative2= fct_relevel(rdnbr_qualitative2, 
                                     levels=c("Other", "High")))

@@

pico <- read_csv("coulter_env_modified.csv") %>% 
  mutate(rdnbr_qualitative = case_when(
    rdnbr_2007 < 69 ~ "None",
    rdnbr_2007 < 315 ~ "Low",
    rdnbr_2007 < 640 ~ "Moderate",
    rdnbr_2007 > 640 ~ "High")) %>%
  mutate(rdnbr_qualitative = as.factor(rdnbr_qualitative),
         rdnbr_qualitative = fct_relevel(rdnbr_qualitative, 
                                         levels=c("High", "Moderate", "Low", "None"))) %>% 
  na.omit(pico) %>%  # 56 stands have NA values (probs all very small stands)
  mutate(rdnbr_qualitative2 = case_when(
    rdnbr_qualitative %in% "High" ~ "High",
    rdnbr_qualitative %in% c("Moderate", "Low", "None") ~ "Other")) %>% 
  mutate(rdnbr_qualitative2 = as.factor(rdnbr_qualitative2)) %>% 
  mutate(rdnbr_qualitative2= fct_relevel(rdnbr_qualitative2, 
                                     levels=c("Other", "High"))) %>% 
  mutate(aspect_transformed = (1-cos((2*pi*aspect_30m)/360))/2) %>% 
  rename_all(funs(str_replace(., "_30m", ""))) %>% 
  mutate(dnbr_qualitative = case_when(
    dnbr_2007 < 0.1 ~ "unburned",
    dnbr_2007 < 0.27 ~ "low",
    dnbr_2007 < 0.66 ~ "moderate",
    dnbr_2007 > 0.66 ~ "high")) %>%
  mutate(dnbr_qualitative = as.factor(dnbr_qualitative),
         dnbr_qualitative = fct_relevel(dnbr_qualitative, levels=c("high", "moderate", "low", "unburned")))

@@

@@

### Sample Size for each mortality class
```{r}
#mcf -------------------------------------------------------------------
mortality_count <- mcf %>% 
  group_by(mortality_qual) %>% 
  count() %>% 
  rename(manual_mortality=mortality_qual)

tc_count <- mcf %>% 
  group_by(tc_mortality) %>% 
  count() # 99 stands with type conversion, 636 with no type conversion

rdnbr_mcf <- mcf %>% 
  group_by(rdnbr_qualitative) %>% 
  count() %>% 
  rename(mcf_rdnbr = rdnbr_qualitative)

dnbr_mcf <- mcf %>% 
  group_by(dnbr_qualitative) %>% 
  count() %>% 
  rename(mcf_dnbr = dnbr_qualitative)

#coulter ------------------------------------------------------------------- 
rdnbr_pico <- pico %>% 
  group_by(rdnbr_qualitative) %>% 
  count() %>% 
  rename(pico_rdnbr =rdnbr_qualitative)

dnbr_pico <- pico %>% 
  group_by(dnbr_qualitative) %>% 
  count() %>% 
  rename(pico_dnbr = dnbr_qualitative)

# all -------------------------------------------------------------------
all_sample_size <- cbind(mortality_count, rdnbr_mcf, dnbr_mcf, rdnbr_pico, dnbr_pico)
```
Yah, dnbr is not a good index to use. Rdnbr is better.

### mortality v rdnbr 
```{r}
rdnbr_qual_n <- mcf %>% 
  group_by(rdnbr_qualitative) %>% 
  count() %>% 
  mutate(mort_type = "rdnbr") %>% 
  rename(mortality_qual = rdnbr_qualitative)

mortality_n <- mcf %>% 
  group_by(mortality_qual) %>% 
  count() %>% 
  mutate(mort_type = "manual")

ggplot(mortality_count, aes(x = mortality_qual, y = n, fill = mort_type)) + geom_col(position = "dodge") +
  geom_text(aes(label=n), position = position_dodge(width=0.9), vjust=-0.5) +
  scale_fill_discrete(name = "How was mortality determined")

@@

rfm <- randomForest(mortality_qual ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + vpd_jan + aspect_transformed + ndvi, 
                    data = mcf, 
                    sampsize = c('None' = 158, 'Low' = 158, 'Moderate' = 158, 'High' = 158), 
                    ntree=1500, 
                    importance=T,
                    replace = T)
# OOB and importance
rfm
importance(rfm)
#plot to visualize variable importance
varImpPlot(rfm)

@@

ALE
```{r}
#sub data
mcf_ale <- mcf %>% 
  select(mortality_qual, rain, tmax, TPI_five, dem_10m_zaca, flow_accumulation, sr, slope, TWI, cwd, vpd_aug, vpd_jan) # the predictor$new code had issues with using a df with NA values even thought the variabels in the RF do not have NA values.

@@

rfm <- randomForest(two_class_mortality_mod ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + vpd_jan + aspect_transformed, 
                    data = mcf_2_class_mod, 
                    sampsize = c('Other' = 158,'High' = 158),
                    ntree=1500, 
                    importance=T,
                    replace = T)
# OOB and importance
rfm
importance(rfm)
#plot to visualize variable importance
varImpPlot(rfm)

@@

ALE
```{r}
#sub data
mcf_ale <- mcf_2_class_mod %>% 
  select(two_class_mortality_mod, rain, tmax, TPI_five, dem_10m_zaca, flow_accumulation, sr, slope, TWI, cwd, vpd_aug, vpd_jan) # the predictor$new code had issues with using a df with NA values even thought the variabels in the RF do not have NA values.

@@

#vpd_jan
other <- plot(FeatureEffect$new(other_predictor, feature = "vpd_jan")) + ggtitle("Other")
high <- plot(FeatureEffect$new(high_predictor, feature = "vpd_jan")) + ggtitle("High")
ggarrange(other, high)

@@

rfm <- randomForest(rdnbr_qualitative2 ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + vpd_jan + aspect_transformed, 
                    data = mcf, 
                    sampsize = c('High' = 150, 'Other' = 150), 
                    ntree=1500, 
                    importance=T,
                    replace = T)
# OOB and importance
rfm
importance(rfm)
#plot to visualize variable importance
varImpPlot(rfm)

@@

@@

@@

>>>>

mcf <- read.csv('Zaca MCF variables for model.csv') %>% 
  mutate(mortality_qual= fct_relevel(mortality_qual, levels=c("High", "Moderate", "Low", "None"))) %>% 
  mutate(dnbr_qualitative = case_when(
    dnbr_2007 < 0.1 ~ "None",
    dnbr_2007 < 0.27 ~ "Low",
    dnbr_2007 < 0.66 ~ "Moderate",
    dnbr_2007 > 0.66 ~ "High")) %>%
  mutate(dnbr_qualitative = as.factor(dnbr_qualitative),
         dnbr_qualitative = fct_relevel(dnbr_qualitative, levels=c("High", "Moderate", "Low", "None"))) %>%
  mutate(aspect_transformed = (1-cos((2*pi*aspect)/360))/2) %>% 
  mutate(three_class_mortality = case_when(
    mortality_qual %in% "High" ~ "High",
    mortality_qual %in% "Moderate" ~ "Moderate",
    mortality_qual %in% c("Low", "None") ~ "Low")) %>% 
  mutate(three_class_mortality = as.factor(three_class_mortality)) %>% 
  mutate(three_class_mortality= fct_relevel(three_class_mortality, 
                                     levels=c("Low", "Moderate", "High"))) %>% 
  mutate(two_class_mortality = case_when(
    mortality_qual %in% "High" ~ "High",
    mortality_qual %in% c("Moderate", "Low", "None") ~ "Other")) %>% 
  mutate(two_class_mortality = as.factor(two_class_mortality)) %>% 
  mutate(two_class_mortality= fct_relevel(two_class_mortality, 
                                     levels=c("Other", "High"))) %>% 
  mutate(tc_mortality = as.factor(tc_mortality)) %>% 
  mutate(rdnbr_qualitative = case_when(
    rdnbr_2007 < 69 ~ "None",
    rdnbr_2007 < 315 ~ "Low",
    rdnbr_2007 < 640 ~ "Moderate",
    rdnbr_2007 > 640 ~ "High")) %>%
  mutate(rdnbr_qualitative = as.factor(rdnbr_qualitative),
         rdnbr_qualitative = fct_relevel(rdnbr_qualitative, 
                                         levels=c("High", "Moderate", "Low", "None"))) %>% 
    mutate(rdnbr_qualitative2 = case_when(
    rdnbr_qualitative %in% "High" ~ "High",
    rdnbr_qualitative %in% c("Moderate", "Low", "None") ~ "Other")) %>% 
  mutate(rdnbr_qualitative2 = as.factor(rdnbr_qualitative2)) %>% 
  mutate(rdnbr_qualitative2= fct_relevel(rdnbr_qualitative2, 
                                     levels=c("Other", "High")))

@@

pico <- read_csv("coulter_env_modified.csv") %>% 
  mutate(rdnbr_qualitative = case_when(
    rdnbr_2007 < 69 ~ "None",
    rdnbr_2007 < 315 ~ "Low",
    rdnbr_2007 < 640 ~ "Moderate",
    rdnbr_2007 > 640 ~ "High")) %>%
  mutate(rdnbr_qualitative = as.factor(rdnbr_qualitative),
         rdnbr_qualitative = fct_relevel(rdnbr_qualitative, 
                                         levels=c("High", "Moderate", "Low", "None"))) %>%
  dplyr::select(-Mortality, -type_converted, -ndvi_2006) %>% 
  na.omit(pico) %>%  # 56 stands have NA values (probs all very small stands)
  #filter(Mortality != "bigcone") %>% # 60 stands actually have big cones not pico
  mutate(rdnbr_qualitative2 = case_when(
    rdnbr_qualitative %in% "High" ~ "High",
    rdnbr_qualitative %in% c("Moderate", "Low", "None") ~ "Other")) %>% 
  mutate(rdnbr_qualitative2 = as.factor(rdnbr_qualitative2)) %>% 
  mutate(rdnbr_qualitative2= fct_relevel(rdnbr_qualitative2, 
                                     levels=c("Other", "High"))) %>% 
  mutate(aspect_transformed = (1-cos((2*pi*aspect_30m)/360))/2) %>% 
  rename_all(funs(str_replace(., "_30m", ""))) %>% 
  mutate(dnbr_qualitative = case_when(
    dnbr_2007 < 0.1 ~ "None",
    dnbr_2007 < 0.27 ~ "Low",
    dnbr_2007 < 0.66 ~ "Moderate",
    dnbr_2007 > 0.66 ~ "High")) %>%
  mutate(dnbr_qualitative = as.factor(dnbr_qualitative),
         dnbr_qualitative = fct_relevel(dnbr_qualitative, levels=c("High", "Moderate", "Low", "None")))

@@

Need to remove the bigcone stands from the pico dataset but having a little trouble atm.

@@

### Sample Size for each mortality class
```{r}
#mcf -------------------------------------------------------------------
mortality_count <- mcf %>% 
  group_by(mortality_qual) %>% 
  count() %>% 
  #rename(manual_mortality=mortality_qual) %>% 
  mutate(mortality_type = 'manual')

rdnbr_mcf <- mcf %>% 
  group_by(rdnbr_qualitative) %>% 
  count() %>% 
  rename(mortality_qual = rdnbr_qualitative) %>% 
  mutate(mortality_type = 'rdnbr')

dnbr_mcf <- mcf %>% 
  group_by(dnbr_qualitative) %>% 
  count() %>% 
  rename(mortality_qual = dnbr_qualitative) %>% 
  mutate(mortality_type = 'dnbr')

mcf_sample_size <- rbind(mortality_count, rdnbr_mcf, dnbr_mcf)

ggplot(mcf_sample_size, aes(x = mortality_qual, y = n, fill = mortality_type)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label=n), position = position_dodge(width=0.9), vjust=-0.5) +
  scale_fill_discrete(name = "How was mortality determined")

# mcf type conversison ------------------------------------------------------------------- 
tc_count <- mcf %>% 
  group_by(tc_mortality) %>% 
  count() # 99 stands with type conversion, 636 with no type conversion

#coulter ------------------------------------------------------------------- 
rdnbr_pico <- pico %>% 
  group_by(rdnbr_qualitative) %>% 
  count() %>% 
  rename(mortality_qual =rdnbr_qualitative) %>% 
  mutate(mortality_type = 'rdnbr')

dnbr_pico <- pico %>% 
  group_by(dnbr_qualitative) %>% 
  count() %>% 
  rename(mortality_qual = dnbr_qualitative)%>% 
  mutate(mortality_type = 'dnbr')

pico_sample_size <- rbind(rdnbr_pico, dnbr_pico)

ggplot(pico_sample_size, aes(x = mortality_qual, y = n, fill = mortality_type)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label=n), position = position_dodge(width=0.9), vjust=-0.5) +
  scale_fill_discrete(name = "How was mortality determined")

```
MCF: Yah, dnbr is not a good index to use. Rdnbr is better at least when it comes to high severity and high mortality. Both dnbr and rdnbr over estimate moderate adn low mortlity and severely underestiamte stands with no mortality. Bc of this, we could make the assumption that pico stands with high rdnbr also have high mortality.

@@

rfm <- randomForest(mortality_qual ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + vpd_jan + aspect_transformed + ndvi + ndmi_2006 + rdnbr_qualitative, 
                    data = mcf, 
                    sampsize = c('None' = 158, 'Low' = 158, 'Moderate' = 158, 'High' = 158), 
                    ntree=1500, 
                    importance=T,
                    replace = T)
# OOB and importance
rfm
importance(rfm)
#plot to visualize variable importance
varImpPlot(rfm)

plot_fi(rf_feat_importance(rfm, mcf))

@@

ALE
```{r}
#sub data
mcf_ale <- mcf %>% 
  select(mortality_qual, rain, tmax, TPI_five, dem_10m_zaca, flow_accumulation, sr, slope, TWI, cwd, vpd_aug, vpd_jan, ndmi_2006, aspect_transformed, ndvi, rdnbr_qualitative) # the predictor$new code had issues with using a df with NA values even thought the variabels in the RF do not have NA values.

@@

rfm <- randomForest(two_class_mortality_mod ~ rain + tmax + TPI_five + flow_accumulation + sr + slope + TWI  + aspect_transformed + ndvi + ndmi_2006, 
                    data = mcf_2_class_mod, 
                    sampsize = c('Other' = 158,'High' = 158),
                    ntree=1500, 
                    importance=T,
                    replace = T)
# OOB and importance
rfm
importance(rfm)
#plot to visualize variable importance
varImpPlot(rfm)

}
```
removed vpd_aug and cwd (rain had greater importance than precip and vpd_jan had greater importance than vpd_aug). .Variable importance got worse, so removed vpd_jan bc tmax became more important more often. NOW the variable iportance is a lot more stable! and lo and behold tmax is more important than slope. removed elevation and replaced it with vpd_jan and variabe importance got cray cray again. Current variables result in very stable importance values!

@@

ALE
```{r}
#sub data
mcf_ale <- mcf_2_class_mod %>% 
  select(two_class_mortality_mod, rain, tmax, TPI_five, dem_10m_zaca, flow_accumulation, sr, slope, TWI, cwd, vpd_aug, vpd_jan, aspect_transformed, ndvi, ndmi_2006) # the predictor$new code had issues with using a df with NA values even thought the variabels in the RF do not have NA values.

@@

#vpd_jan
other <- plot(FeatureEffect$new(other_predictor, feature = "ndvi")) + ggtitle("Other")
high <- plot(FeatureEffect$new(high_predictor, feature = "ndvi")) + ggtitle("High")
ggarrange(other, high)

@@

rfm <- randomForest(rdnbr_qualitative2 ~ rain + tmax + TPI_five + dem_10m_zaca + flow_accumulation + sr + slope + TWI + cwd + vpd_aug + vpd_jan + aspect_transformed + ndvi + ndmi_2006, 
                    data = mcf, 
                    sampsize = c('High' = 150, 'Other' = 150), 
                    ntree=1500, 
                    importance=T,
                    replace = T)
# OOB and importance
rfm
importance(rfm)
#plot to visualize variable importance
varImpPlot(rfm)

@@

ALE
```{r}
#sub data
mcf_ale <- mcf %>% 
  select(rdnbr_qualitative2, rain, tmax, TPI_five, dem_10m_zaca, flow_accumulation, sr, slope, TWI, cwd, vpd_aug, vpd_jan, aspect_transformed, ndvi, ndmi_2006) # the predictor$new code had issues with using a df with NA values even thought the variabels in the RF do not have NA values.

@@

```
interesting. for rdnbr classification, the ale plots show different trends for ndvi. .However, this might change as the explanatory variables change

@@

# categorical variable 
none <-  plot(FeatureEffect$new(none_predictor, feature = "rdnbr_qualitative")) + ggtitle("None")
low <- plot(FeatureEffect$new(low_predictor, feature = "rdnbr_qualitative")) + ggtitle("Low")
moderate <- plot(FeatureEffect$new(moderate_predictor, feature = "rdnbr_qualitative")) + ggtitle("Moderate")
high <- plot(FeatureEffect$new(high_predictor, feature = "rdnbr_qualitative")) + ggtitle("High")
ggarrange(none, low, moderate, high)

# plot 2 variables
none <-  plot(FeatureEffect$new(none_predictor, feature = c("tmax", "rain"))) + ggtitle("None")
low <- plot(FeatureEffect$new(low_predictor, feature = c("tmax", "rain"))) + ggtitle("Low")
moderate <- plot(FeatureEffect$new(moderate_predictor, feature = c("tmax", "rain"))) + ggtitle("Moderate")
high <- plot(FeatureEffect$new(high_predictor, feature = c("tmax", "rain"))) + ggtitle("High")
ggarrange(none, low, moderate, high)
